Community Analysis

### 2022-07-05 R/09-community-analysis.Rmd
combine all edits to incorporate the figures and analyses into the manuscript

---
title: "Community Analysis"
author: "Andrew Morris"
date: "9/15/2021"
output: html_document
editor_options: 
  chunk_output_type: inline
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = TRUE)
knitr::opts_knit$set(root.dir=normalizePath('../'))

@@

```{r initialize_packages}
library(tidyverse)
library(phyloseq)
library(decontam)
library(microViz)
library(knitr)
library(corncob)
library(DivNet)
library(magrittr)
library(vegan)
library(ggpubr)
library(ggthemes)
library(broom)

@@

```{r import-data}
load(file = "Output/phyloseq.RData")
ps <-
  ps %>% 
  ps_mutate(pas_by_treat = paste0(passage, treat))
ps_melt <-
  ps %>% 
  ps_melt()
```

@@

```{r plot-depth}

depth_data <-
  ps %>% 
  ps_melt %>% 
  group_by(Sample, passage, treat) %>% 
  summarize(depth = sum(Abundance))

summary(aov(depth ~ passage * treat, data = depth_data))
depth_data %>% 
  ggplot(aes(x = interaction(treat, passage), y = depth)) +
  theme_tufte() +
  geom_rangeframe() +
  # coord_cartesian(clip="off") +
  geom_tufteboxplot()

@@

There is no systematic difference in sequencing depth/library size between treatments. It does appear that sequencing depth decreases between passage 2 and 5.

```{r plot-richness}

plot_richness(ps, x = "pas_by_treat", color="treat", measures=c("Observed", "Shannon", "Simpson"), title = "Alpha Diversity ASV") +
    scale_color_manual(name = "Treatment", 
                         labels = c('Neutral', 'Positive'),
                         values = c('gray40', 'darkorange2'))

@@

Richness decreases with passage, but seems unchanged between 5n and 5p.
Initial richness appears similar between the treatments.

# Hypothesis Testing

```{r richness}

# Test effect of passage number on richness using all samples
richness_soil <- ps %>% breakaway
plot(richness_soil, physeq=ps, color="passage")
summary(richness_soil) %>% as_tibble

meta <- ps %>%
  sample_data %>%
  as_tibble %>%
  mutate("sample_names" = ps %>% sample_names )

combined_richness <- meta %>%
  left_join(summary(richness_soil),
            by = "sample_names")

bt_passage <- betta(chats = combined_richness$estimate,
                      ses = combined_richness$error,
                      X = model.matrix(~passage, data = combined_richness))
bt_passage$table

# Test effect of treatment on richness in passage 5

ps_5 <- ps_filter(ps, passage == "5")
richness_soil <- ps_5 %>% breakaway
plot(richness_soil, physeq=ps_5, color="treat")
summary(richness_soil) %>% as_tibble

meta <- ps_5 %>%
  sample_data %>%
  as_tibble %>%
  mutate("sample_names" = ps_5 %>% sample_names )

combined_richness <- meta %>%
  left_join(summary(richness_soil),
            by = "sample_names")

bt_treat5 <- betta(chats = combined_richness$estimate,
                      ses = combined_richness$error,
                      X = model.matrix(~treat, data = combined_richness))
bt_treat5$table

# Test effect of CH4 on richness

bt_ch4 <- betta(chats = combined_richness$estimate,
                      ses = combined_richness$error,
                      X = model.matrix(~ch4, data = combined_richness))
bt_ch4$table

save(bt_passage, bt_treat5, bt_ch4, file = 'Output/richness_models.RData')

@@

Test confirms the above observation: Richness decreases between Passage 2
and Passage 5. No difference in richness by treatment in passage 5 or by
methane flux.

@@

@@

# Replace unknown and missing taxa
ps_fixed <-
ps %>%
 tax_fix(
  min_length = 4,
  unknowns = c(""),
  sep = " ", anon_unique = TRUE,
  suffix_rank = "classified"
 )

# Recode treatments as binary 1/0
ps_fixed <- 
  ps_fixed %>% 
  ps_mutate(treat_bin = factor(ifelse(treat=='p', 1,0))) %>% 
  ps_mutate(passage_bin = factor(ifelse(passage==5, 1,0))) %>% 
  ps_mutate(treat_pas = interaction(treat_bin, passage_bin))

ps_fixed
```

```{r calc-beta, results=FALSE}

ord_calc_aitch <-
  ps_fixed %>% 
  tax_transform(trans = 'clr') %>% 
  ord_calc(method = "PCA")

ord_calc_bray <-
  ps_fixed %>%
  dist_calc("bray") %>%
  ord_calc(method = "NMDS")

ord_calc_jacc <-
  ps_fixed %>%
  dist_calc("jaccard", binary = TRUE) %>%
  ord_calc(method = "PCoA")

@@

## Bray-Curtis and Jaccard

```{r plot-beta, results=FALSE, fig.width = 8, fig.height = 3}
bray <-
  ord_calc_bray %>% 
  ord_plot(color = "treat") +
  theme_tufte() +
  geom_rangeframe() +
  coord_cartesian(clip="off") +
  scale_color_manual(name = "Treatment", 
                       labels = c('Neutral', 'Positive'),
                       values = c('gray40', 'darkorange2')) +
  labs(title = "Bray-Curtis")

jacc <- 
  ord_calc_jacc %>% 
  ord_plot(color = "treat") +
  theme_tufte() +
  geom_rangeframe() +
  coord_fixed(3.9/48.3) +
  # coord_cartesian(clip="off") +
  scale_color_manual(name = "Treatment", 
                       labels = c('Neutral', 'Positive'),
                       values = c('gray40', 'darkorange2')) +
  labs(title = "Jaccard")

ggarrange(bray, jacc)

@@

## Aitchison

```{r}
plot_beta(ps_fixed)
```

```{r beta-test}

ps_clr <- microbiome::transform(ps, "clr")

fit_bin_jacc <- adonis2(otu_table(ps) ~ passage * treat, data = as(sample_data(ps), 'data.frame'), method = "jaccard", binary = TRUE)

fit_aitch <- adonis2(otu_table(ps_clr) ~ passage * treat, data = as(sample_data(ps_clr), 'data.frame'), method = "euclidean")

fit_adonis <- rbind(tidy(fit_aitch), tidy(fit_bin_jacc))
fit_adonis$term <- c('Passage', 'Treatment', 'Pass:Treat', 'Residual', 'Total',
      'Passage', 'Treatment', 'Pass:Treat', 'Residual', 'Total')
save(fit_adonis, file = 'Output/beta_model.RData')

############

clr_dist_matrix <- phyloseq::distance(ps_clr, method = "euclidean")

dispr <- vegan::betadisper(clr_dist_matrix, phyloseq::sample_data(ps_clr)$passage)
disper_test <- permutest(dispr)

dispr <- vegan::betadisper(clr_dist_matrix, phyloseq::sample_data(ps_clr)$treat)
disper_test <- permutest(dispr)

dist_aitch5 <-
  ps %>% 
  ps_filter(passage == 5) %>% 
  dist_calc(dist = "aitchison")
dist_permanova(dist_aitch5, "treat")
groups <- factor(c(rep(1,12), rep(2,12)), labels = c("N","P"))

dist_aitch5 %>% 
  dist_get %>% 
  vegan::betadisper(groups) %>% 
  permutest()

dist_aitch2 <-
  ps %>% 
  ps_filter(passage == 2) %>% 
  dist_calc(dist = "aitchison")
dist_permanova(dist_aitch5, "treat")
groups <- factor(c(rep(1,12), rep(2,10)), labels = c("N","P"))

dist_aitch2 %>% 
  dist_get %>% 
  vegan::betadisper(groups) %>% 
  permutest()

@@

There is a significant main effect and interaction of passage and treatment on
taxonomic dissimilarity of the soil microbiome. In addition, there is no difference
in dispersion between treatments, but there is between passage2 and 5.

>>>>

---
title: "Community Analysis"
author: "Andrew Morris"
date: "6/28/2022"
output: 
  html_document:
    toc: true
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}

knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = TRUE)
knitr::opts_knit$set(root.dir=normalizePath('../'))
set.seed(1337)

@@

```{r initialize_packages}

library(phyloseq)
library(knitr)
library(magrittr)
library(vegan)
library(ggpubr)
library(ggthemes)
library(broom)
library(ggtext)
library(ALDEx2)
library(corncob)
library(ANCOMBC)
library(tidyverse)

@@

```{r import-data}

aes_data <- read_tsv('Output/community_data.tsv')

```{r fix-treatment}

aes_data <-
  aes_data %>% 
  mutate(passage_treat = treat) %>% 
  separate(passage_treat, c('passage', "treat"), sep = 2, remove = FALSE) %>% 
  select(sample, passage_treat, passage, treat, selected, ch4, asv, count, Kingdom:Species, seq)

```{r fix-taxonomy}

# There are "Unknown Family" taxa with Genus and Class, but Order is Incertae Sedis.
# I'm just going to remove everything below Class and call it "Unclassified"

fill_taxonomy <- function(tab) {
    if(is.na(tab[[1]])) tab[[1]] = "Unclassified"
    for(i in 2:length(tab)) {
      if(is.na(tab[[i]])){
        if(grepl("unclassified", tab[[i-1]])) tab[[i]] = tab[[i-1]]
        else tab[[i]] = paste(tab[[i-1]], "unclassified", sep = "_")
      }
    }
    tab
}

aes_data <-
  aes_data %>% 
  nest(data = -c(asv, Kingdom:Species)) %>% 
  nest(taxa = Kingdom:Species) %>% 
  mutate(taxa = map(taxa, ~fill_taxonomy(.x))) %>% 
  unnest(taxa) %>% 
  unnest(data)

@@

```{r n-seqs}

aes_data %>% 
  group_by(sample) %>% 
  summarize(total = sum(count)) %>% 
  ggplot(aes(x = total)) +
  geom_histogram() +
  theme_classic() +
  labs(title = "Distribution sequencing depths",
       x = "Library size",
       y = "Frequency")

seq_range <-
  aes_data %>% 
  group_by(sample) %>% 
  summarize(total = sum(count)) %>% 
  summarize(seq_range = range(total)) %$% 
  seq_range

a_analysis <- 
  aes_data %>%
  count(sample, count) %>%
  nest(data = -sample) %>%
  mutate(n_seqs = map_dbl(data, ~sum(.x$count * .x$n)),
         sobs = map_dbl(data, ~sum(.x$n)),
         rare = map_dbl(data, ~rarefy(rep(.x$count, .x$n), sample=seq_range[1]))) %>%
  select(-data)

a_analysis %>%
  select(sample, n_seqs, sobs, rare) %>%
  pivot_longer(-c(sample, n_seqs)) %>%
  ggplot(aes(x=n_seqs, y=value, color=name)) +
  geom_point() +
  geom_smooth(se=FALSE) +
  coord_cartesian(ylim=c(0, 6000)) +
  theme_classic() +
  labs(x = "Sequencing Depth", y = "Alpha Diversity")

a_analysis %>%
  select(sample, n_seqs, sobs, rare) %>%
  mutate(treat = substr(sample, 1, 3)) %>% 
  select(-sample) %>% 
  group_by(treat) %>% 
  pivot_longer(-c(treat)) %>%
  ggplot(aes(x=treat, y=value)) +
  facet_wrap(~ name, nrow = 3, scales = "free") +
  geom_boxplot() +
  geom_jitter() +
  theme_classic() +
  labs(title = "AES", x = "Treatment", y = "Value")

@@

This is the number of reads (n_seqs), the rarefied richness (rare), and 
the observed richness (sobs for "species observed") for the four treatments:
generations 2 and 5 and selection treatments neutral and positive. As you can
see, there were more sequences from G2 than G5 and also greater richness
in G2 than G5. However, that richness difference is robust to rarefying
to the level of the shallowest sample. There doesn't seem to be a systemic
difference in sequencing depth or richness between the neutral and positive 
treatments in either generation.

### Plot

```{r alpha-diversity-plot}

# aes_data %>% 
#   group_by(sample) %>% 
#   summarize(n_seqs = sum(count)) %>% 
#   summarize(min = min(n_seqs))

aes_df <-
  aes_data %>% 
  select(sample, asv, count) %>%
  pivot_wider(names_from = "asv", values_from = "count", values_fill = 0) %>% 
  as.data.frame()

rownames(aes_df) <- aes_df$sample
aes_df <- aes_df[, -1]

rare <- 
  rarefy(aes_df, seq_range[1]) %>% 
  as_tibble(rownames = "sample") %>% 
  rename(rare = value)

aes_data %>% 
  inner_join(rare, by = 'sample') %>% 
  select(sample, passage_treat, rare) %>% 
  distinct %>% 
  ggplot(aes(x = passage_treat, y = rare)) +
  geom_boxplot() +
  geom_jitter() +
  labs(x = "Passage x Treatment", y = "Rarefied Richness") +
  theme_bw() +
  theme(
    plot.background = element_blank(),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank())

@@

### Hypothesis Test

```{r alpha-diversity-test}

rare_data <-
  aes_data %>% 
  select(sample, passage_treat, passage, treat, ch4) %>% 
  distinct %>% 
  inner_join(rare, by = "sample")

kt <- kruskal.test(rare ~ passage_treat, data = rare_data)
kt <- tidy(kt)

wt <- pairwise.wilcox.test(rare_data$rare, 
                           rare_data$passage_treat, 
                           p.adjust.method = "BH")
wt <- tidy(wt)

rare_median <-
  rare_data %>% 
  group_by(passage) %>% 
  summarize(median = median(rare),
            IQR = quart(rare)[2] - quart(rare)[1])

@@

There is an effect of passage x treatment on richness (Kruskal-Wallace test: chi-squared = 
`r kt[1,1]`, df = `r kt[1,3]`, p = `r kt[1, 2]`). The pairwise test shows that
G2 is different from G5, but N and P are not different in either generation
(Wilcoxon/Mann-Whitney Test).

kable(wt)

```{r alpha-diversity-ch4-test}

s <- cor.test(~ ch4 + rare, data = rare_data, method = "spearman", alternative = "two.sided") %>% 
  tidy

s5 <-
  rare_data %>% 
  filter(passage == "G5") %>% 
  cor.test(~ ch4 + rare, data = ., method = "spearman", alternative = "two.sided")
save(kt, wt, rare_median, s, file = "Output/richness_models.RData")

A Spearman's rank correlation test shows no correlation between
methane oxidation rate and richness (Spearman's rho = `r s[1,1]`, p = `r s[1,3]`).

@@

```{r bray-curtis-nmds, messages=FALSE}

dist <- vegdist(aes_df, method="bray")

if(file.exists("Output/rare_dist_bray.rds")) {
  rare_dist_bray <- 
    readRDS("Output/rare_dist_bray.rds")
} else {
  rare_dist_bray <- avgdist(aes_df, dmethod="bray", sample=seq_range[1])
  rare_dist_bray %>% 
    saveRDS("Output/rare_dist_bray.rds")
}

nmds <- metaMDS(rare_dist_bray)

nmds_scores <-
  scores(nmds, display = "sites") %>% 
  as_tibble(rownames = "sample")

nmds_data <-
  aes_data %>% 
  select(sample, passage, treat) %>% 
  distinct %>% 
  inner_join(nmds_scores, by = "sample")

centroids <-
  nmds_data %>% 
  group_by(passage, treat) %>% 
  summarize(NMDS1 = mean(NMDS1),
            NMDS2 = mean(NMDS2),
            .groups = "drop")
p_beta <-
  nmds_data %>%   
  ggplot(aes(x = NMDS1, y = NMDS2, shape = passage, color = treat)) +
  geom_point() +
  stat_ellipse(show.legend = FALSE) +
  coord_fixed() +
  scale_shape_discrete(name = "Passage") +
  scale_color_manual(name = "Treatment",
                       labels = c('Neutral', 'Positive'),
                       values = c('gray40', 'darkorange2')) #+
  # theme_bw() +
  # theme(
  #   plot.background = element_blank(),
  #   panel.grid.major = element_blank(),
  #   panel.grid.minor = element_blank())

save(p_beta, file = "Output/beta_fig.RData")

ggsave('Manuscript/fig3.jpg', p_beta, height = 2.5, width = 4)

```{r permanova-passage-treatment, warnings=FALSE}

dist_df <-
  rare_dist_bray %>% 
  as.matrix %>% 
  as_tibble(rownames = "sample")

meta_dist <-
  aes_data %>% 
  select(sample, passage_treat, passage, treat, ch4) %>% 
  distinct %>% 
  inner_join(dist_df, by = "sample")

permanova <- adonis2(rare_dist_bray ~ passage * treat, data = meta_dist, permutations = 999)

betadisper(rare_dist_bray, meta_dist$passage_treat) %>% 
  permutest()

tidy_permanova <- 
  tidy(permanova)

save(tidy_permanova, file = "Output/beta_model.RData")

There is a significant difference of centroid between passage, treatment, and
their interaction. Testing a difference of dispersion using betadisper
shows no difference in variance between the four groups.

# Distance-based RDA Methane

```{r full-dbrda}

meta_dist %>% 
  select(passage_treat, ch4) %>% 
  mutate(log_ch4 = log_methane(ch4)) %>% 
  ggplot(aes(x = passage_treat, y = log_ch4)) +
  geom_boxplot() +
  theme_classic() +
  labs(x = "Passage x Treatment", y = "Log10 Methane")

dbrda_ch4 <- dbrda(rare_dist_bray ~ rank(ch4), 
                      data = meta_dist, 
                      na.action = "na.omit",
                      add = TRUE)

dbrda_ch4 %>% 
  permutest(permutations = 9999)

percent_explained <- 100 * eigenvals(dbrda_ch4) / sum(eigenvals(dbrda_ch4))

dbrda_scores <-
  scores(dbrda_ch4, display = "sites") %>% 
  as_tibble(rownames = "sample")

dbrda_data <-
  aes_data %>% 
  select(sample, passage, treat, ch4) %>% 
  distinct %>% 
  inner_join(dbrda_scores, by = "sample")

dbrda_data %>%   
  ggplot(aes(x = dbRDA1, y = MDS1, shape = passage, color = treat)) +
  geom_point() +
  stat_ellipse(show.legend = FALSE) +
  coord_fixed() +
  scale_shape_discrete(name = "Passage") +
  scale_color_manual(name = "Treatment",
                       labels = c('Neutral', 'Positive'),
                       values = c('gray40', 'darkorange2')) +
  theme_bw() +
  theme(
    plot.background = element_blank(),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank())

```{r pre-dbrda-data-subsets}

# aes_data %>% 
#   filter(treat == "P") %>% 
#   group_by(asv) %>% 
#   mutate(total = sum(count)) %>% 
#   filter(total != 0) %>% 
#   ungroup() %>% 
#   select(-total) %>% 
#   group_by(sample) %>% 
#   summarize(n_seqs = sum(count)) %>% 
#   summarize(min = min(n_seqs))

treatP_df <-
  aes_data %>%  
  filter(treat == "P") %>% 
  group_by(asv) %>% 
  mutate(total = sum(count)) %>% 
  filter(total != 0) %>% 
  ungroup() %>% 
  select(-total) %>% 
  select(sample, asv, count) %>%
  pivot_wider(names_from = "asv", values_from = "count", values_fill = 0) %>% 
  as.data.frame()

rownames(treatP_df) <- treatP_df$sample
treatP_df <- treatP_df[, -1]

# aes_data %>% 
#   filter(passage == "G5") %>% 
#   group_by(asv) %>% 
#   mutate(total = sum(count)) %>% 
#   filter(total != 0) %>% 
#   ungroup() %>% 
#   select(-total) %>% 
#   group_by(sample) %>% 
#   summarize(n_seqs = sum(count)) %>% 
#   summarize(min = min(n_seqs))

passage5_df <-
  aes_data %>%  
  filter(passage == "G5") %>% 
  group_by(asv) %>% 
  mutate(total = sum(count)) %>% 
  filter(total != 0) %>% 
  ungroup() %>% 
  select(-total) %>% 
  select(sample, asv, count) %>%
  pivot_wider(names_from = "asv", values_from = "count", values_fill = 0) %>% 
  as.data.frame()

rownames(passage5_df) <- passage5_df$sample
passage5_df <- passage5_df[, -1]

```{r treat-dbrda}

if(file.exists("Output/treat_rare_dist_bray.rds")) {
  treat_rare_dist_bray <- 
    readRDS("Output/treat_rare_dist_bray.rds")
} else {
treat_rare_dist_bray <- avgdist(treatP_df, dmethod="bray", sample=seq_range[1])
  treat_rare_dist_bray %>% 
    saveRDS("Output/treat_rare_dist_bray.rds")
}

treat_dist_df <-
  treat_rare_dist_bray %>% 
  as.matrix %>% 
  as_tibble(rownames = "sample")

treat_meta_dist <-
  aes_data %>% 
  select(sample, passage_treat, passage, treat, ch4) %>% 
  distinct %>% 
  inner_join(treat_dist_df, by = "sample")

treat_dbrda_ch4 <- dbrda(treat_rare_dist_bray ~ rank(ch4), 
                      data = treat_meta_dist, 
                      na.action = "na.omit",
                      add = TRUE)

treat_dbrda_ch4 %>% 
  permutest(permutations = 9999)

percent_explained <- 100 * eigenvals(treat_dbrda_ch4) / sum(eigenvals(treat_dbrda_ch4))

treat_dbrda_scores <-
  scores(treat_dbrda_ch4, display = "sites") %>% 
  as_tibble(rownames = "sample")

treat_dbrda_data <-
  aes_data %>% 
  select(sample, passage, treat, ch4) %>% 
  distinct %>% 
  inner_join(treat_dbrda_scores, by = "sample")

treat_dbrda_data %>%   
  ggplot(aes(x = dbRDA1, y = MDS1, shape = passage, color = treat)) +
  geom_point() +
  stat_ellipse(show.legend = FALSE) +
  coord_fixed() +
  scale_shape_discrete(name = "Passage") +
  scale_color_manual(name = "Treatment",
                       labels = c('Positive'),
                       values = c('darkorange2')) +
  theme_bw() +
  theme(
    plot.background = element_blank(),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank())

```{r passage-dbrda}
if(file.exists("Output/passage_rare_dist_bray.rds")) {
  passage_rare_dist_bray <- 
    readRDS("Output/passage_rare_dist_bray.rds")
} else {
passage_rare_dist_bray <- avgdist(passage5_df, dmethod="bray", sample=seq_range[1])
  passage_rare_dist_bray %>% 
    saveRDS("Output/passage_rare_dist_bray.rds")
}

passage_dist_df <-
  passage_rare_dist_bray %>% 
  as.matrix %>% 
  as_tibble(rownames = "sample")

passage_meta_dist <-
  aes_data %>% 
  select(sample, passage_treat, passage, treat, ch4) %>% 
  distinct %>% 
  inner_join(passage_dist_df, by = "sample")

passage_dbrda_ch4 <- dbrda(passage_rare_dist_bray ~ rank(ch4), 
                      data = passage_meta_dist, 
                      na.action = "na.omit",
                      add = TRUE)

passage_dbrda_ch4 %>% 
  permutest(permutations = 9999)

passage_percent_explained <- 100 * eigenvals(passage_dbrda_ch4) / sum(eigenvals(passage_dbrda_ch4))

passage_dbrda_scores <-
  scores(passage_dbrda_ch4, display = "sites") %>% 
  as_tibble(rownames = "sample")

passage_dbrda_data <-
  aes_data %>% 
  select(sample, passage, treat, ch4) %>% 
  distinct %>% 
  inner_join(passage_dbrda_scores, by = "sample")

passage_dbrda_data %>%   
  ggplot(aes(x = dbRDA1, y = MDS1, shape = passage, color = treat)) +
  geom_point() +
  stat_ellipse(show.legend = FALSE) +
  coord_fixed() +
  scale_shape_discrete(name = "Passage") +
  scale_color_manual(name = "Treatment",
                       labels = c('Neutral', 'Positive'),
                       values = c('gray40', 'darkorange2')) +
  theme_bw() +
  theme(
    plot.background = element_blank(),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank())

The overall dbRDA is significant and rank of methane explains 5% of the variation.
Subsetting the Positive treatment in both G2 and G5, it is not significant: no
correlation between dissimilarity and rank of methane between G2 and G5. This
makes sense, because most of that variation would be due to selection in the jar.
However, subsetting passage G5 and comparing N and P, 10% of dissimilarity is
explained by rank of CH4.

## Unique to each treatment

```{r}
g5n <- 
  aes_data %>% 
  filter(passage_treat == "G5N") %>% 
  select(Family) %>% 
  distinct() %$%
  Family

g5p <- 
  aes_data %>% 
  filter(passage_treat == "G5P") %>% 
  select(Family) %>% 
  distinct() %$%
  Family

p_not_n <- g5p[!g5p %in% g5n]

n_not_p <- g5n[!g5n %in% g5p]

n_not_p_prevalence <-
  aes_data %>% 
  filter(passage_treat == "G5N") %>% 
  filter(Family %in% n_not_p) %>% 
  select(sample, Family, count) %>% 
  mutate(prevalence = as.numeric(count > 0)) %>% 
  select(-count) %>% 
  group_by(Family) %>% 
  summarize(prevalence = sum(prevalence))

n_not_p_prevalence %>% 
  ggplot(aes(x = prevalence)) +
  geom_histogram() +
  scale_x_continuous(breaks = 1:12) +
  theme_classic() +
  labs(title = "Unique to Positive")

# n_not_p_prevalence %>%
#   filter(prevalence == 1) %>% 
#   nrow

# n_not_p_prevalence %>%
#   filter(prevalence == 2) %>% 
#   nrow
# n_not_p_prevalence %>%
#   filter(prevalence > 2) %$%
#   Family

p_not_n_prevalence <-
  aes_data %>% 
  filter(passage_treat == "G5P") %>% 
  filter(Family %in% p_not_n) %>% 
  select(sample, Family, count) %>% 
  mutate(prevalence = as.numeric(count > 0)) %>% 
  select(-count) %>% 
  group_by(Family) %>% 
  summarize(prevalence = sum(prevalence))

p_not_n_prevalence %>% 
  ggplot(aes(x = prevalence)) +
  geom_histogram() +
  scale_x_continuous(breaks = 1:12) +
  theme_classic() +
  labs(title = "Unique to Neutral")
# 
# p_not_n_prevalence %>%
#   filter(prevalence == 1) %>% 
#   nrow
# 
# p_not_n_prevalence %>%
#   filter(prevalence == 2) %>% 
#   nrow
p_not_n_over_6 <- p_not_n_prevalence %>%
  filter(prevalence > 6)

First, I looked at which Families are unique to each treatment in passage 5.
There were `r length(n_not_p)` families unique to N and `r length(p_not_n)`
unique to P. However, of the  `r length(n_not_p)` unique to P, 
`r n_not_p_prevalence %>% filter(prevalence == 1) %>% nrow` had a prevalence
of 1 and `r n_not_p_prevalence %>% filter(prevalence == 2) %>% nrow` had
a prevalence of 2 leaving only a single family, which had a prevalence of
`r n_not_p_prevalence %>% filter(prevalence > 2) %$% prevalence`, which was `r n_not_p_prevalence %>% filter(prevalence > 2) %$% Family`.
Of the `r length(p_not_n)` Families unique to P, 2 had prevalences greater than
6, which were `r p_not_n_over_6$Family[1]` with a prevalence of
`r p_not_n_over_6$prevalence[1]` and `r p_not_n_over_6$Family[2]` with
a prevalence of `r p_not_n_over_6$prevalence[2]`.

## Setup Differential Abundance Data

```{r setup-differential-abundance-data}

aes_G5 <-
  aes_data %>% 
  filter(passage == "G5")

aes_family_G5 <-
  aes_G5 %>% 
  select(asv, sample, treat, count, Family) %>%
  group_by(sample, Family) %>% 
  mutate(total = sum(count)) %>% 
  ungroup() %>% 
  select(sample, treat, Family, total) %>% 
  distinct()

prevalent_families <- 
  aes_family_G5 %>% 
  group_by(Family, treat) %>% 
  summarize(prev = sum(total > 0), .groups = "drop") %>% 
  pivot_wider(names_from = "treat", values_from = "prev", values_fill = 0) %>% 
  filter(N/12 > 0.1 & P/12 > 0.1) %$%
  Family

otu_table <- 
  aes_family_G5 %>% 
  filter(Family %in% prevalent_families) %>% 
  select(sample, Family, total) %>% 
  pivot_wider(names_from = sample, values_from = total, values_fill = 0) %>% 
  as.data.frame()

rownames(otu_table) <- otu_table[,1]
otu_table <- otu_table[,-1]

sample_data <- aes_family_G5 %>% 
  select(sample, treat) %>% 
  distinct() %>% 
  as.data.frame()

rownames(sample_data) <- sample_data[,1]

## Corncob

@@

da_analysis <- differentialTest(formula = ~ treat,
                                phi.formula = ~ 1,
                                formula_null = ~ 1,
                                phi.formula_null = ~ 1,
                                test = "Wald", 
                                boot = FALSE,
                                data = otu_table,
                                sample_data = sample_data,
                                fdr_cutoff = 0.05, B = 100,
                                taxa_are_rows = TRUE)

dt_data <- plot(da_analysis, 
                # level = c("Family"), 
                data_only = TRUE)

dt_plot <- function(dt_data) {
  ggplot(dt_data, aes(x = x, xmin = xmin, xmax = xmax, y = reorder(taxa, x))) +
    geom_pointrange() +
    geom_vline(xintercept = 0) +
    labs(x = "Estimate", y = NULL) +
    coord_cartesian(clip="off") +
    # theme(panel.border = element_blank()) +
    theme_classic() +
      theme(axis.text.y = element_markdown())

}

dt_data %>%
  mutate(taxa = str_replace(taxa, "(.*)", "*\\1*"),
         taxa = str_replace(taxa, "\\*(.*)_unclassified\\*",
                                "Unclassified *\\1*")) %>% 
  dt_plot()

@@

## ANCOM

# Setup phyloseq object
aes_ps <- phyloseq(otu_table(otu_table, taxa_are_rows=TRUE), sample_data(sample_data))

if(file.exists("Output/ancom_out.rds")) {
  out <- 
    readRDS("Output/ancom_out.rds")
} else {
  # Run Ancom
  out = ancom(phyloseq = aes_ps,  
              p_adj_method = "BH",
              prv_cut = 0.1, 
              lib_cut = 0, 
              main_var = "treat",
              rand_formula = NULL, 
              lme_control = NULL,
              struc_zero = TRUE, 
              neg_lb = FALSE, 
              alpha = 0.05, 
              n_cl = 1)  
  out %>% 
    saveRDS("Output/ancom_out.rds")
}

res = out$res
q_val = out$q_data
beta_val = out$beta_data
# Only consider the effect sizes with the corresponding q-value less than alpha
beta_val = beta_val * (q_val < 0.05) 
# Choose the maximum of beta's as the effect size
beta_pos = apply(abs(beta_val), 2, which.max) 
beta_max = vapply(seq_along(beta_pos), function(i) beta_val[beta_pos[i], i],
                  FUN.VALUE = double(1))
# Number of taxa except structural zeros
n_taxa = ifelse(is.null(out$zero_ind), 
                nrow(feature_table), 
                sum(apply(out$zero_ind, 1, sum) == 0))
# Cutoff values for declaring differentially abundant taxa
cut_off = 0.7 * (n_taxa - 1)

df_fig_w = res %>%
  dplyr::mutate(beta = beta_max,
                direct = case_when(
                  detected_0.7 == TRUE & beta > 0 ~ "Positive",
                  detected_0.7 == TRUE & beta <= 0 ~ "Negative",
                  TRUE ~ "Not Significant"
                  )) %>%
  dplyr::arrange(W)
df_fig_w$taxon_id = factor(df_fig_w$taxon_id, levels = df_fig_w$taxon_id)
df_fig_w$W = replace(df_fig_w$W, is.infinite(df_fig_w$W), n_taxa - 1)
df_fig_w$direct = factor(df_fig_w$direct, 
                     levels = c("Negative", "Positive", "Not Significant"))

p_w = df_fig_w %>%
  ggplot(aes(x = taxon_id, y = W, color = direct)) +
  geom_point(size = 2, alpha = 0.6) +
  labs(x = "Taxon", y = "W") +
  scale_color_discrete(name = NULL) + 
  geom_hline(yintercept = cut_off, linetype = "dotted", 
             color = "blue", size = 1.5) +
  geom_text(aes(x = 2, y = cut_off + 0.5, label = "W[0.7]"), 
            size = 5, vjust = -0.5, hjust = 0, color = "orange", parse = TRUE) +
  theme_bw() +
  theme(axis.text.x = element_blank(),
        axis.ticks.x = element_blank(),
        panel.grid.major = element_blank())
p_w

@@

## Aldex2

conds <- colnames(otu_table) %>% 
  substr(3, 3)

if(file.exists("Output/aldex_x_all.rds")) {
  x.all <- 
    readRDS("Output/aldex_x_all.rds")
} else {
  x.all <- aldex(otu_table, 
                 conds, mc.samples=1000, test="t", effect=TRUE,
                 include.sample.summary=FALSE, denom="all", verbose=FALSE, paired.test=FALSE)
  x.all %>% 
    saveRDS("Output/aldex_x_all.rds")
}

par(mfrow=c(1,2))
aldex.plot(x.all, type="MA", test="welch", xlab="Log-ratio abundance",
    ylab="Difference")
aldex.plot(x.all, type="MW", test="welch", xlab="Dispersion",
    ylab="Difference")

par(mfrow=c(1,2))
plot(x.all$effect, x.all$we.ep, log="y", cex=0.7, col=rgb(0,0,1,0.2),
  pch=19, xlab="Effect size", ylab="P value", main="Effect size plot")
points(x.all$effect, x.all$we.eBH, cex=0.7, col=rgb(1,0,0,0.2),
  pch=19)
abline(h=0.05, lty=2, col="grey")
legend(15,1, legend=c("P value", "BH-adjusted"), pch=19, col=c("blue", "red"))

plot(x.all$diff.btw, x.all$we.ep, log="y", cex=0.7, col=rgb(0,0,1,0.2),
  pch=19, xlab="Difference", ylab="P value", main="Volcano plot")
points(x.all$diff.btw, x.all$we.eBH, cex=0.7, col=rgb(1,0,0,0.2),
  pch=19)
abline(h=0.05, lty=2, col="grey")

x.all_effect_1 <-
  x.all %>% 
  filter(abs(effect) > 1) %>% 
  as_tibble(rownames = "Family")

# x.all_effect_1 %>% 
#   # inner_join(taxa_table, by = "asv")
#   ggplot(aes(y = reorder(Family, effect), x = effect)) +
#   geom_point()

# x.all_effect_1 %>% 
#   select(Family, effect)

@@

## Comparison

### All Significant

# CORNCOB
corncob_taxa <- dt_data$taxa

# ANCOM
ancom_taxa <- as.character(df_fig_w[df_fig_w$detected_0.7, ]$taxon_id)

# ALDEX2
aldex_taxa <- x.all_effect_1$Family

at_least_one_taxa <-
	aes_family_G5 %>%
	group_by(sample) %>%
	mutate(rel_abund = (total / sum(total)) * 100) %>%
	filter(Family %in% ancom_taxa | Family %in% aldex_taxa | Family %in% corncob_taxa) %>%
	mutate(Family = str_replace(Family, "(.*)", "*\\1*"),
	       Family = str_replace(Family, "\\*(.*)_unclassified\\*",
				    "Unclassified *\\1*"))

at_least_one_iqm <- 
	at_least_one_taxa %>%
	group_by(treat, Family) %>%
	summarize(median = median(rel_abund), 
		  first_quartile = quart(rel_abund)[1],
		  third_quartile = quart(rel_abund)[2], .groups = "drop")

sort_order <-
	at_least_one_taxa %>%
	group_by(Family) %>%
	summarize(median = median(rel_abund)) %>%
	arrange(median)

at_least_one_taxa$Family <- factor(at_least_one_taxa$Family, levels = sort_order$Family)
